Ignition of Fuel Beds by Firebrands

### 2022-05-22 manuscript4.tex
Added CFD plot to manuscript 4

\renewcommand{\TheAddress}{
\textbf{Status: In Preparation}
}

@@

                    \draw [<-] (3.5, 0.1) -- (3, 0.5) node[left] {\scriptsize Heater};
                    \draw [<-] (3.5, 1.55) -- (3, 1.55) node[left] {\scriptsize Load Cell};
                    \draw [<-] (3.75, 2.5) -- (3, 2.5) node[left] {\scriptsize Lowering Arm};
                    \draw[->]         (0.1, 0.6) -- (0.75, 0.6);
                    \draw[->]         (0.1, 1.2) -- (0.75, 1.2);
                    \node[right] at (-0.75, 1.5) {\scriptsize Inlet};
                    \draw[->]         (0.1, 1.8) -- (0.75, 1.8);
                    \draw[->]         (0.1, 2.4) -- (0.75, 2.4);
                    \draw[draw=black, dashed] (3.336, 0) rectangle (4.54, 0.34);
                    \draw[->]         (4.3, 1) -- (4.0, 0.34);
                    \node[right, align=left] at (4.3, 1) {\scriptsize Computational\\ \scriptsize Domain};
                    \fill[draw=red, fill=red] (3.43, 0.01) circle (0.0635);
                    \fill[draw=red, fill=red] (3.74, 0.01) circle (0.0635);
                \end{tikzpicture}
                }
            \caption{Diagram of the experimental wind tunnel apparatus. Air flows through the wind tunnel from left to right. The dashed region represents the domain subset used for computational efforts.}
            \label{fig:multiHeaterApparatus}
        \end{figure}

@@

\section{Results and Discussion}
    The flaming ignition or non-ignition result of each test is shown in Figure~\ref{fig:multi_heater_hot_vs_ambient}. The markers show the result of each test and the the curves show the logistic regression for each of the heater configurations as defined in Table~\ref{tab:multiHeaterConfig}. The shaded regions around the curves represent the 95\% confidence intervals of each regression. The top plot in Figure~\ref{fig:multi_heater_hot_vs_ambient} shows configurations where the downstream heater is actively heated and the upstream heater is at ambient conditions. The bottom plot in Figure~\ref{fig:multi_heater_hot_vs_ambient} shows results for configurations where both heaters are actively heated to the same temperature.
        \begin{figure}[hpbt]
            \centering
            \includegraphics[width=0.75\columnwidth]{Figures/multi_heater_plot.png}
            \caption{Ignition or no ignition outcomes of tests with different heater configurations. The circular markers indicate outcomes of individual tests and the curves represent logistic regressions of each test series. The shaded region shows the 95\% confidence interval for each regression.}
            \label{fig:multi_heater_hot_vs_ambient}
        \end{figure}
    Table~\ref{tab:multiFiftyTemp} shows the heater temperatures estimated to result in a 50\% ignition probability from the logistic regressions shown in Figure~\ref{fig:multi_heater_hot_vs_ambient} as well  Temperature values for the corresponding wind speed and orientation from the single heater study in Chapter~\ref{part:manuscript2} are shown in the column titled "Single" for comparison. 
        \rowcolors{2}{gray!25}{white}
        \begin{table}[hpbt]
            \normalsize
            \caption{Heater temperature required to achieve 50\% ignition probability for the configurations tested with 95\% confidence intervals shown in parentheses. The columns "Unheated" and "Heated" refer to the temperature of the upstream heater. The single heater column shows the values from the study in Chapter~\ref{part:manuscript2} for a single heater.}
            \centering
            \begin{tabular}{ccrrr}
                \rowcolor{gray!50}
               U\textsubscript{bulk} (\si{\meter\per\second}) & Spacing (D) & Unheated (\si{\celsius})& Heated (\si{\celsius}) & Single (\si{\celsius})\\
                \hline
                0.5  & 1 & 750 (644, 850) & 533 (527, 541) & 571\\
                0.5  & 5 & 578 (565, 583) & 601 (597, 604) & 571\\
                5.8  & 1 & 423 (417, 429) & 406 (397, 416) & 399\\
                5.8  & 5 & 438 (421, 454) & 396 (387, 407) & 399
            \end{tabular}
            \label{tab:multiFiftyTemp}
        \end{table}

    Considering first the cases where both heaters are actively heated (Figure~\ref{fig:multi_heater_hot_vs_ambient} bottom panel) there are three observations of note. First, in both the one and five heater diameter spacing cases an increase in wind from \textit{0.5\si{\meter\per\second}} to \texit{5.8\si{\meter\per\second}} resulted in a lower temperature threshold for ignition. The 50\% ignition threshold decreased 24\% (127\si{\celsius}) for the one diameter spacing and 34\% (205\si{\celsius}) for the five diameter spacing configuration. These results are similar to the 35\% decrease in the ignition threshold for the single heater results presented in Chapter~\ref{part:manuscript2}. The reduction in ignition temperature is attributed to an increased residence time of pyrolyzates near the heater in recirculation zones enabling ignition at lower temperatures. The second observation for the 0.5\si{\meter\per\second} wind speed tests is that decreasing the heater spacing from five heater diameters to one heater diameter resulted in an 11\% decrease in the temperature required for 50\% ignition probability. The shift in ignition probabilities is attributed to increased heat transfer between the closely spaced heaters. Anecdotally, the shift in ignition probability observed by moving the heaters closer together was accompanied by a shift in ignition mode. As the heater temperature was reduced below the 50\% ignition threshold for a single heater, ignition was observed after the smoldering fronts between the heaters merged and ignition occurred in the plume of pyrolyzates between the heaters. Noting that the ignition threshold for the one diameter spaced heaters is lower than that of the single heater the lower ignition threshold is attributed to the increased heat transfer resulting from two heat sources. It is not clear from these results what specific mechanism is the driving factor. Possibilities include increased temperature of pyrolyzates departing between the heaters due to to enhanced radiative and convective heat transfer, preheating of air as it flows over the upstream heater before mixing, increased mass flux of pyrolyzates from the fuel bed, or a combination of effects. In contrast to the differences in ignition at a wind speed of 0.5\si{\meter\per\second} the differences in ignition probability were not statistically significant at 5.8\si{\meter\per\second}. This suggests that the enhanced heat transfer between the cartridge heaters is not influential in windy conditions. Anecdotally, in cases under wind ignitions were to occur on the upstream side of the upstream cartridge heater suggesting that ignition is more favorable in the recirculation zones on the upstream side of the heater even with the addition of heat and pyrolysis.

@@

    More analysis to come.

>>>>

\renewcommand{\TheAddress}{
\textbf{Status: In Preparation}\\
Target Journals: Fire Safety Science, Fire Safety Journal, or Fire Technology
}

@@

                    \draw [<-] (3.5, 1.55) -- (3, 1.55) node[left] {\scriptsize Load Cell};
                    \draw [<-] (3.75, 2.5) -- (3, 2.5) node[left] {\scriptsize Lowering Arm};
                    \draw[->]         (0.1, 0.6) -- (0.75, 0.6);
                    \draw[->]         (0.1, 1.2) -- (0.75, 1.2);
                    \node[right] at (-0.75, 1.5) {\scriptsize Inlet};
                    \draw[->]         (0.1, 1.8) -- (0.75, 1.8);
                    \draw[->]         (0.1, 2.4) -- (0.75, 2.4);
                    \draw[draw=black, dashed] (2.84 , 0) rectangle (4.34, 0.7);
                    \draw[->]         (4.3, 1) -- (4.0, 0.34);
                    \node[right, align=left] at (4.3, 1) {\scriptsize Computational\\ \scriptsize Domain};
                    \fill[draw=red, fill=red] (3.43, 0.01) circle (0.0635);
                    \fill[draw=red, fill=red] (3.74, 0.01) circle (0.0635);
                    \draw [<-] (3.4, 0.1) -- (2.85, 0.5) node[left] {\scriptsize Heaters};
                \end{tikzpicture}
                }
            \caption{Diagram of the experimental wind tunnel apparatus. Air flows through the wind tunnel from left to right. The dashed region represents the domain subset used for computational efforts.}
            \label{fig:multiHeaterApparatus}
        \end{figure}

@@

    A simplified model was implemented to obtain further insight into the fluid mechanics and heat transfer processes that influence the trends observed in the experimental efforts. The model was implemented in three parts. First, a thermal model of the fuel bed was used to estimate the mass of the fuel bed above the pyrolysis temperature and the average temperature. Second, the mass of the fuel bed material converted from solid fuel to gaseous pyrolysis products was estimated. In the third and final step the fluid flow and heat transfer between the heaters, air, and released pyrolyzates was modeled. The modelling efforts were conducted for a 10\si{\second} interval to ensure proper characterization of the flow field and maintain an accurate representation of the fuel bed. As mass is lost from the fuel bed due to pyrolysis the contact area between the heater and the fuel bed changes as does the surface shape of the fuel bed. It was observed during experiments that the contact between the heat and the fuel bed and the shape of the fuel bed itself began to deviate significantly from the model domain at approximately 10\si{\second}. Considerations of changes in contact area and changing geometry due to pyrolysis is out of the scope of this work. Nonetheless, the treatment of each configuration equally and within the limited time frame is still anticipated to produce insight into the influence of the thermal and fluid processes on ignition.

    The thermal model of the fuel bed was conducted using OpenFOAM~\cite{Foundation2020}. Figure~\ref{fig:multiHeaterThermal} shows the computational domain of the fuel bed. The thermal diffusivity of the fuel bed materials was estimated using correlations from literature for the thermal conductivity~\cite{bergman2011fundamentals} and published values of specific heat~\cite{Parker1986}. Equation~\ref{eqn:estk4} shows the the correlation for effective thermal conductivity where $\epsilon$, as defined in Equation~\ref{eqn:epsilon4}, is the proportion of the fuel bed that is air. The bulk density of the fuel bed material ($\rho_{bed}$) and the solid wood ($\rho_{solid}$) were obtained from experimental samples. 
        \begin{equation}
            k_{eff} = \frac{1}{2} \left(\frac{1}{\left( 1 - \epsilon \right)/k_{solid} + \epsilon/k_{air}} +  \epsilon k_{air} + \left(1-\epsilon \right) k_{solid} \right)
            \label{eqn:estk4}
        \end{equation}
        \begin{equation}
            \epsilon = 1- \frac{\rho_{solid}}{\rho_{bed}}
            \label{eqn:epsilon4}
        \end{equation}
    The interface between the fuel bed and air and the radial edge of the domain were treated as insulated. The centerline was treated as a symmetric boundary condition. The insulated radial edge and symmetric boundary condition are consistent with boundary conditions in the actual experiment since the 25\si{\milli\meter} distance ensures no temperature change during 10\si{\second} of heating modeled and the heater temperature has been observed to be circumferentially consistent with infrared imaging. The insulated interface with the air does not account for radiation and convective heat losses from the surface of the fuel bed as it is heated. However, the magnitude of the heat losses from the fuel bed is estimated to be significantly smaller in magnitude than the heat transfer from the heater. Reactions, mass loss due to pyrolysis, and changes in contact area due to changing fuel bed geometry are not considered in the temperature analysis. While the aforementioned assumptions and limitations reduce the accuracy of the model in comparison to the experimental conditions, the temperature distributions still provide insights into differences in the mass of the fuel bed material that undergoes pyrolysis and provides insight to better interpret the experimental results. 
    \begin{figure}
        \centering
        \resizebox{0.375\textwidth}{!}{%
          \begin{tikzpicture}
            \node at (0,0) (origin) {};
            %\draw (origin) rectangle (topright);
            \fill[pattern=north west lines, pattern color= brown, draw=black] (3.18mm, 0) -- (25mm, 0) arc (0:-90:25mm) -- (0, -3.18mm) arc (270:0:3.18mm ) -- cycle ;

            \node (text) at (12mm, -12mm) {Fuel Bed};
            \fill [draw = white,  inner color=red, outer color=white ] (0mm, 0mm) circle (10mm);
            \fill [draw=white, fill=white] (0,0) -- (10mm, 0) arc(0:270:10mm) -- (0,0);
            \filldraw [fill=white, draw=red, thick] (0, 0) circle (3.18mm);

            \fill[pattern=north east lines, pattern color=black!90] (-2mm,-3.18mm) rectangle (0mm,-27mm) node[rotate=-90, pos=0.5] {};
            %\fill[pattern=north east lines, pattern color=black!90] (0mm,-2mm) rectangle (27mm,0mm) node[rotate=-90, pos=0.5] {};
           \fill[pattern=crosshatch] (25mm, 0) -- (27mm, 0) arc (0:-90:27mm) -- (0, -25mm) arc(-90:0:25mm) ;
            \draw[-stealth,decorate,decoration={snake,amplitude=1pt,pre length=0pt,post length=2pt}] (3.5mm, -6mm) -- ++(1mm, 9mm);
            \draw[-stealth,decorate,decoration={snake,amplitude=1pt,pre length=0pt,post length=2pt}] (6mm, -5mm) -- ++(1mm, 7mm);
            \draw [|-|] (-5mm, 0mm) -- (-5mm, -25mm) node[pos=0.5, left] {25mm} ;
             \draw[thick, dashed] (3.18mm, 0) -- (10mm, 0);
            \draw (0mm, -3.18mm) -- (0mm, -10mm);
            \node[draw, rectangle, fill=white, draw=white] (p) at (8mm, 10mm) {Pyrolysis Gases};
	        \draw[->, thick] (p.south) -- (6mm, 2mm);
	        \node[draw, rectangle, fill=white, draw=white, anchor=west] (p) at (12mm, 4mm) {T=220\si{\celsius}};
	        \draw[->, thick] (p.west) -- (10mm, 1mm);
          \end{tikzpicture}
          }
        \caption{Diagram of the computational domain where black lines indicate domain boundaries, and red lines are boundaries defined by the heater. The arrows denote flow of pyrolysis products from the fuel bed into the air above the fuel bed.}
        \label{fig:multiHeaterThermal}
    \end{figure}

    The velocity of pyrolyzates entering the fluid domain from the fuel bed was determined in a two step process. First the mass of material in the fuel bed undergoing pyrolysis was estimated. All material over a temperature of 220\si{\celsius} was considered to undergo pyrolysis. This temperature was chosen as 220\si{\celsius} corresponds to the minimum temperature for the pyrolysis of hemicellulose and is considered the onset of pyrolysis~\cite{Yang2007a}. The amount of mass estimated to depart the fuel bed was determined to be the proportion of fuel bed material that converted to gas phase products when reacted at the average temperature of the material above the pyrolysis threshold from the thermal model. Reactions were conducted using the BioPOx mechanism~\cite{Dhahak2019} and mass was assumed to depart the fuel bed if it existed in the gas phase. Species were considered to be gas phase if they were present in the Bio1412 mechanism~\cite{Ranzi2001, Ranzi2008}, since the Bio1412 mechanism considers the gas phase reaction of pyrolysis products. The reaction calculations were conducted in Cantera~\cite{Goodwin2020}. The Cantera calculations were 0D which necessitated using the average temperatures of the fuel bed for the pyrolysis calculations rather than spatially resolved temperatures. Mass flux of the gases exiting the fuel bed was determined by averaging the total mass of gaseous pyrolysis products produced over the 10\si{\second} interval. The exit velocity was then determined as the mass flux through the distance between the the heater edge and the pyrolysis threshold as shown by the dotted line in Figure~\ref{fig:multiHeaterThermal}. The calculated velocity and injection area was then used as an input for calculations conducted in the flow field.

    A 2D RANS model of the flow above the fuel bed and around the heaters was implemented in OpenFOAM. The area of the computational domain with respect to the experimental apparatus is shown as the dashed line in Figure~\ref{fig:multiHeaterApparatus}. The 2D slice of the domain modeled was considered to pass through the center of the heaters and corresponds to the location of the highest temperature of the heater which also typically coincides with the ignition location. Turbulence parameters were determined from measurements taken of the experimental apparatus with a TSI IFA-300 hot wire anemometer. The k-$\epsilon$ turbulence model was used with values of 1.4$\cdot$10\textsuperscript{-4}\si{\square\meter\per\square\second} and 9.5$\cdot$10\textsuperscript{-6}\si{\square\meter\per\cubic\second} for k and $\epsilon$ respectively. Calculations were conducted for 10\si{\second} and were then averaged to observed differences in temperature and velocity characteristics between the different configurations.

\section{Results and Discussion}
    The flaming ignition or non-ignition result of each test is shown in Figure~\ref{fig:multi_heater_hot_vs_ambient}. The markers show the result of each test and the the curves show the logistic regression for each of the heater configurations as defined in Table~\ref{tab:multiHeaterConfig}. The shaded regions around the curves represent the 95\% confidence intervals of each regression. The top plot in Figure~\ref{fig:multi_heater_hot_vs_ambient} shows configurations where the downstream heater is actively heated and the upstream heater is at ambient conditions. The bottom plot in Figure~\ref{fig:multi_heater_hot_vs_ambient} shows results for configurations where both heaters are actively heated to the same temperature.
        \begin{figure}[hpbt]
            \centering
            \includegraphics[width=0.75\columnwidth]{Figures/multi_heater_plot.png}
            \caption{Ignition or no ignition outcomes of tests with different heater configurations. The circular markers indicate outcomes of individual tests and the curves represent logistic regressions of each test series. The shaded region shows the 95\% confidence interval for each regression.}
            \label{fig:multi_heater_hot_vs_ambient}
        \end{figure}
    Table~\ref{tab:multiFiftyTemp} shows the heater temperatures estimated to result in a 50\% ignition probability from the logistic regressions shown in Figure~\ref{fig:multi_heater_hot_vs_ambient} as well  Temperature values for the corresponding wind speed and orientation from the single heater study in Chapter~\ref{part:manuscript2} are shown in the column titled "Single" for comparison. 
        \rowcolors{2}{gray!25}{white}
        \begin{table}[hpbt]
            \normalsize
            \caption{Heater temperature required to achieve 50\% ignition probability for the configurations tested with 95\% confidence intervals shown in parentheses. The column "Unheated" refers to cases where the upstream heater was not temperature controlled and "Heated" refers to cases where both upstream and downstream heaters were maintained at the setpoint temperature. The single heater column shows the values from the study in Chapter~\ref{part:manuscript2} for a single heater.}
            \centering
            \begin{tabular}{ccrrr}
                \rowcolor{gray!50}
               U\textsubscript{bulk} (\si{\meter\per\second}) & Spacing (D) & Unheated (\si{\celsius})& Heated (\si{\celsius}) & Single (\si{\celsius})\\
                \hline
                0.5  & 1 & 750 (644, 850) & 533 (527, 541) & 571\\
                0.5  & 5 & 578 (565, 583) & 601 (597, 604) & 571\\
                5.8  & 1 & 423 (417, 429) & 406 (397, 416) & 399\\
                5.8  & 5 & 438 (421, 454) & 396 (387, 407) & 399
            \end{tabular}
            \label{tab:multiFiftyTemp}
        \end{table}

    Considering first the cases where both heaters are actively heated (Figure~\ref{fig:multi_heater_hot_vs_ambient} bottom panel and Table~\ref{tab:multiFiftyTemp} "Heated" column) there are three observations of note. First, in both the one and five heater diameter spacing cases an increase in wind from \textit{0.5\si{\meter\per\second}} to \textit{5.8\si{\meter\per\second}} resulted in a lower temperature threshold for ignition. The 50\% ignition threshold decreased 24\% (127\si{\celsius}) for the one diameter spacing and 34\% (205\si{\celsius}) for the five diameter spacing configuration. These results are similar to the 35\% decrease in the ignition threshold for the single heater results presented in Chapter~\ref{part:manuscript2}. The reduction in ignition temperature is attributed to an increased residence time of pyrolyzates near the heater in recirculation zones enabling ignition at lower temperatures. The second observation for the 0.5\si{\meter\per\second} wind speed tests is that decreasing the heater spacing from five heater diameters to one heater diameter resulted in an 11\% decrease in the temperature required for 50\% ignition probability. The shift in ignition probabilities is attributed to increased heat transfer between the closely spaced heaters. Anecdotally, the shift in ignition probability observed by moving the heaters closer together was accompanied by a shift in ignition mode. As the heater temperature was reduced below the 50\% ignition threshold for a single heater, ignition was observed only after the smoldering fronts between the heaters merged and ignition occurred in the plume of pyrolyzates that established between the heaters. The lower threshold for ignition when both heaters were heated and only one diameter apart is attributed to increased increased heat transfer to the pyrolyzates as they depart the fuel bed. It is not clear from these results what specific mechanism is the driving factor. Possibilities include increased temperature of pyrolyzates departing between the heaters due to to enhanced radiative and convective heat transfer, preheating of air as it flows over the upstream heater before mixing, increased mass flux of pyrolyzates from the fuel bed, or a combination of effects. In contrast to the differences in ignition at a wind speed of 0.5\si{\meter\per\second} the differences in ignition probability were not statistically significant at 5.8\si{\meter\per\second}. This suggests that the enhanced heat transfer between the cartridge heaters is not influential in windy conditions. Anecdotally, in cases under wind ignitions were to occur on the upstream side of the upstream cartridge heater suggesting that ignition is more favorable in the recirculation zones on the upstream side of the heater even with the addition of heat and pyrolysis.

@@

    Further insight into the influence of heater configuration and wind speed on ignition is gained from the results of the computational efforts. Figure~\ref{fig:multiHeaterCFD} shows the averaged velocity streamlines and temperature distribution in the fluid flow around the heaters for each heater configuration. Air flows from left to right in Figure~\ref{fig:multiHeaterCFD}. The numbers inside the circular heater boundaries correspond to the assigned temperature of the heater during the calculation. The assigned surface temperatures correspond to the heater temperature anticipated to produce a 50\% probability of ignition for the wind speed and heater conditions. Heater boundaries with blue circles inside them correspond to cases where the heater was not heated and acted as an inert object. The arrows indicate the observed location of ignition, when observed. Ignition locations were not observed for every case, however, sufficient ignition locations were recorded to provide insight into controlling parameters.
        \begin{figure}
            \centering
            \includegraphics[width=\columnwidth]{Figures/quarter_circle.pdf}
            \caption{Time averaged velocity streamlines and temperature profiles for each configuration. The numbers inside of the heater boundaries identify surface temperature of the heater. Blue circles indicate that the heater was unheated and acting as an inert object. Arrows typical ignition locations.}
            \label{fig:multiHeaterCFD}
        \end{figure}
